---
title: "Ghana Education Analytics Report"
author: "Samuel Mintah"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
ghana <- read_csv("../data/processed/ghana_education_clean.csv")
```

## Overview

This report explores Ghana's education trends from 2000 to 2023 using
data from the World Bank Education Statistics database.

## Enrollment Trends

Primary enrollment has grown significantly over the past two decades,
while secondary enrollment remains comparatively low.
```{r enrollment-chart}
ghana %>%
  select(year, enrollment_primary, enrollment_secondary) %>%
  pivot_longer(-year, names_to = "level", values_to = "enrollment") %>%
  mutate(level = recode(level,
    "enrollment_primary"   = "Primary",
    "enrollment_secondary" = "Secondary")) %>%
  ggplot(aes(x = year, y = enrollment, colour = level)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2) +
  labs(title = "Ghana School Enrollment Rates (2000–2023)",
       x = "Year", y = "Enrollment Rate (%)", colour = "Level",
       caption = "Source: World Bank WDI") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Gender Gap
```{r gender-chart}
ghana %>%
  select(year, enrollment_primary_f, enrollment_primary_m) %>%
  pivot_longer(-year, names_to = "gender", values_to = "enrollment") %>%
  mutate(gender = recode(gender,
    "enrollment_primary_f" = "Female",
    "enrollment_primary_m" = "Male")) %>%
  ggplot(aes(x = year, y = enrollment, colour = gender)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2) +
  scale_colour_manual(values = c("Female" = "#E91E8C", "Male" = "#1E90FF")) +
  labs(title = "Gender Gap in Primary Enrollment — Ghana",
       x = "Year", y = "Enrollment Rate (%)", colour = "Gender",
       caption = "Source: World Bank WDI") +
  theme_minimal() + theme(legend.position = "bottom")
```

## Out-of-School Children
```{r oos-chart}
ghana %>%
  filter(!is.nan(out_of_school)) %>%
  ggplot(aes(x = year, y = out_of_school)) +
  geom_col(fill = "#E74C3C", alpha = 0.8) +
  labs(title = "Out-of-School Children Rate — Ghana",
       x = "Year", y = "Out-of-School Rate (%)",
       caption = "Source: World Bank WDI") +
  theme_minimal()
```

## Pupil-Teacher Ratio
```{r ptr-chart}
ghana %>%
  filter(!is.nan(pupil_teacher_ratio)) %>%
  ggplot(aes(x = year, y = pupil_teacher_ratio)) +
  geom_line(linewidth = 1.2, colour = "#2ECC71") +
  geom_hline(yintercept = 40, linetype = "dashed", colour = "red") +
  annotate("text", x = 2001, y = 41, label = "UNESCO threshold (40)",
           colour = "red", size = 3) +
  labs(title = "Pupil-Teacher Ratio — Ghana Primary Schools",
       x = "Year", y = "Pupils per Teacher",
       caption = "Source: World Bank WDI") +
  theme_minimal()
```

## Key Findings

- Primary enrollment has improved from **`r round(ghana$enrollment_primary[1], 1)`%** 
in 2000 to over **90%** in recent years
- A persistent gender gap exists, with boys consistently enrolled 
at higher rates than girls
- Out-of-school rates have declined but remain a challenge
- Pupil-teacher ratios have stayed near the UNESCO recommended threshold of 40